\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-07-27 mers-structure.tex
Generalise discussion, AR edits.

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology}

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike simpler ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
% TB: Agree.
structured population models possess the ability to estimate migration times. % So do Markov Jumps.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters %beyond what could be expected of phylogenetic noise %% doesn't make sense to say that recombination is interfering as expected by phylogenetic noise.
(Figure \ref{recombination}A).
And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
%Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases. %% a bit overblown 'extreme caution'?
On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches. %% But why won't recombination be happening in humans? Sure it won't be 2 divergent lineages.

@@

\subsection*{Role of population structure in phylodynamics}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{ABC}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{ABC}).
We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).
Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.

Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods. %% Don't understand this sentence. How is the second clause related to the first?
As computational infrastructure improves \citep{ayres_beagle:_2012}, %it is hopefully only a matter of time before
phylogenetic population structure models will reach the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) will be tractable.

@@

\subsection*{Sequence data as a universal epidemiology tool} %% Drop this paragraph and finish on something better. I think this point has been made already.
As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
%Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}  %% AR - sorry, I hate this sentence - 'fundamental unit of ground truth' WTF?
In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.

@@

\section*{Acknowledgements}
We would like to thank \tbc{XXX, YYY, ZZZ, etc...} for kindly sharing sequence data, and Allison Black for useful discussion and advice.
GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.

>>>>

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface}

@@

\subsection*{Signals of MERS-CoV seasonality}
% Unlike the most simpler %\lmc{simpler?}
% ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
% structured population models possess the ability to estimate migration times. % So do Markov Jumps. \gdc{good point, yeah}
We use the posterior distribution of MERS-CoV introduction times of human clades observed in the MCC tree (Figure \ref{mcc}) recovered from structured coalescent to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters %beyond what could be expected of phylogenetic noise %% doesn't make sense to say that recombination is interfering as expected by phylogenetic noise.
(Figure \ref{recombination}A).
And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Observed departures from strictly clonal evolution suggest that whilst recombination is an issue for inferring MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
%Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases. %% a bit overblown 'extreme caution'?
In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.
% On the other hand, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches. %% But why won't recombination be happening in humans? Sure it won't be 2 divergent lineages.

@@

\subsection*{MERS-CoV epidemiology}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (Figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (Figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (Figure \ref{ABC}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (Figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (Figure \ref{ABC}).
We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (Figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (Figure \ref{ABC}).

\subsection*{Transmissibility differences between zoonoses and pandemics}
Severe Acute Respiratory Syndrome coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.

% Despite higher resolution of inference offered by structured population models, especially when sequence data are plentiful, serious drawbacks exist.
% As the representation of evolution drifts away from a strictly bifurcating tree model, the space over which phylogenetic inference needs to take place becomes increasingly complex.
% Convergence difficulties are well known in rich models, such as ancestral recombination graphs, and multitype trees are no different.
% Even with a moderate number of sequences (N=274) and the smallest possible number of demes (N=2) structured population model mixing is slow and problematic.
% Some convergence issues (Figure \ref{skygrid}) could be attributed to the presence of recombination in MERS-CoV data (Figure \ref{recombination}), but active research into approximations to the structured coalescent \citep{maio_new_2015} seem to acknowledge both the utility and difficulties in applying such methods. %% Don't understand this sentence. How is the second clause related to the first?
% As computational infrastructure improves \citep{ayres_beagle:_2012}, %it is hopefully only a matter of time before
% phylogenetic population structure models will reach the point where large datasets with multiple demes (\textit{e.g.} West African Ebola virus epidemic \citep{dudas_virus_2017}, global influenza virus circulation \citep{bedford_global_2015}) will be tractable.

@@

% \subsection*{Sequence data as a universal epidemiology tool} %% Drop this paragraph and finish on something better. I think this point has been made already.
% As the field of epidemiology enters the age of genomics sequence data are likely to be brought in to reinforce and augment epidemiological investigations.
% At a relatively low price, sequence data offer the combined benefits of diagnostics and typing, whilst also offering insight into a pathogen's evolutionary history.
% With a small amount of metadata, primarily collection dates and features of interest, such as location or host, much can be revealed through phylogenetic clustering of sequences, or absence thereof.
% This was used to great effect at all stages of the Ebola virus epidemic in West Africa, where sequence data early on were used to show continued circulation of the same strain in the region rather than multiple zoonotic introductions \citep{gire_genomic_2014}, to study viral migration at a broad level \citep{park_ebola_2015,ladner_evolution_2015} as healthcare infrastructure was being overwhelmed during the peak, and towards the end where the final transmission chains were tracked and confirmed as such \citep{arias_rapid_2016}.
% %Although sequence data can be collected for multiple reasons and used to address different questions, sequence data are a fundamental unit of ground truth, which allows sequences from wildly different studies to be combined trivially%\lmc{TRIVIALLY???} \gdc{yes, since you literally just put sequences into the same alignment}  %% AR - sorry, I hate this sentence - 'fundamental unit of ground truth' WTF?
% In fact it has often been the combination of data from distinct populations and different time periods that yielded most insight in the past \citep{dudas_virus_2017}.

@@

\section*{Acknowledgements}
We would like to thank \tbc{XXX, YYY, ZZZ, etc...} for kindly sharing sequence data, and Allison Black for useful discussion and advice.
GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
We gratefully acknowledge the contribution of the following scientists for sharing MERS-CoV sequence data before publication:

Ali M. Somily$^{1}$, Mazin Barry$^{1}$, Sarah S. Al Subaie$^{1}$, Abdulaziz A. BinSaeed$^{1}$, Fahad A. Alzamil$^{1}$, Waleed Zaher$^{1}$, Theeb Al Qahtani$^{1}$, Khaldoon Al Jerian$^{1}$, Scott J.N. McNabb$^{2}$, Imad A. Al-Jahdali$^{3}$, Ahmed M. Alotaibi$^{4}$, Nahid A. Batarfi$^{5}$, Matthew Cotten$^{6}$, Simon J. Watson$^{6}$, Spela Binter$^{6}$, Paul Kellam$^{6}$.

$^{1}$College of Medicine, King Saud University, Riyadh, Kingdom of Saudi Arabia \\
$^{2}$Rollins School of Public Health, Emory University, Atlanta, GA, USA \\
$^{3}$Deputy Minister. Ex. General Director King Fahad General Hospital, Jeddah and Occupational and environmental medicine, Um AlQura University, Kingdom of Saudi Arabia \\
$^{4}$Department of Intensive Care, Prince Mohammed bin Abdulaziz Hospital, Riyadh, Kingdom of Saudi Arabia \\
$^{5}$Epidemiology section, Command and Control Center (CCC) Ministry of Health, Jeddah \\
$^{6}$Wellcome Trust Sanger Institute, Hinxton, United Kingdom \\
